Effect of Transmission Type on Miles Per Gallon
===============================================

Executive Summary
-----------------
In this paper we examine the Motor Trends data on the effect of transmission type on
miles per gallon (mpg).  First we determine whether an automatic or manual transmission
is better for mpg.  For this we conducted a t-test for the difference of mean mpg between
cars with automatic or manual transmission.  At $\alpha=0.05$ we find there is a statistically
significant difference in mpg between automatic and manual transmissions.  Second we quantifying the effect of transmission type on mpg.  Using a linear regression model we see that a 
manual transmission can increse the expected mpg by 2.936 when all other factors are
held constant.

Data Analysis
-------------
```{r ,warning=FALSE, echo=FALSE}
library(plyr)
get.formula <- function(x) {
    f <- formula(x)
    paste(f[2],f[1],f[3],sep=" ")
}
cars <- mutate(mtcars,
               vs=factor(vs),
               am=factor(ifelse(am,"manual","automatic"),
                         levels=c("automatic","manual")))

```
Obtain the transmission type (automatic or manual) and miles per gallon rating for cars in the Motor Trend Car Road Test. First we assess whether a manual or automatic transmission is better for mpg.  According to Figure 1 in the Appendix, automatic transmissions provide lower mpg.
```{r ,echo=FALSE, echo=FALSE}
## t test for difference of the mean
auto <- subset(cars,cars$am=="automatic",select=c(mpg))[,1]
man <- subset(cars,cars$am=="manual",select=c(mpg))[,1]
t <- t.test(auto,man)
```
We verify this by conducting a t-test for the difference mean mpg by transmission
type.  At $\alpha=0.05$, we see the t statistic is `r t$statistic` with a p-value 
of `r t$p.value`, which indicates there is a statistically significant difference
between the respective mean values for mpg.

Modeling
--------
We prepare the data for modeling.  For non-categorical attributes, the mean is 
subtracted.  This will allow for realistic interpretation of the intercept term.
```{r ,echo=FALSE}
## prepare data for modeling
cars <- mutate(cars,
               wt=(wt-mean(wt)), qsec=(qsec-mean(qsec)), disp=(disp - mean(disp)),
               hp=(hp - mean(hp)), drat=(drat-mean(drat)), gear=(gear-mean(gear)),
               carb=(carb-mean(carb))
               )
```

```{r, warning=FALSE, echo=FALSE}
##  model relationship
lm1 <- lm(mpg~.,data=cars); lm1.s <- summary(lm1); f<-lm1.s$fstatistic
```
To quantify the effect of transmission type on mpg, we create a linear regression model 
using all the relevant variables: **Model 1 `r get.formula(lm1)`**.  For this model the F-statistic is `r f[1]` with a p-value of `r pf(f[1],f[2],f[3],lower.tail=FALSE)`, which show the model is
significant at the $\alpha=0.05$.  While the model may be significant, the
p-value for the coefficients indicate we cannot reject the null hypothesis that they are zero.
```{r ,echo=FALSE}
lm1.s$coefficients
```
```{r ,message=FALSE,echo=FALSE,results='hide'}
lm2 <- step(lm1, direction="both"); lm2.s <- summary(lm2)
```
To find an alternative model, we use the **step** function to explore other formulations of the regression model.  After running the procedure, the **step** function selected this formulation as the optimal representation: **Model 2 `r get.formula(lm2)`**. For this second model, the F-statistic is `r  lm2.s$fstatistic[1]` with a p-value is `r f<-lm2.s$fstatistic;  pf(f[1],f[2],f[3],lower.tail=FALSE)`  The selected model has these coefficients, which based on the respective p-values are 
statistically significant.
```{r ,echo=FALSE}
lm2.s$coefficients
```
The model's intercept shows the expected mpg is `r round(coef(lm2)[1],3)` for a car with
the average weight and average seconds for a quater mile.  The coefficient for wt indicates
the mpg will decrease by `r round(abs(coef(lm2)[2]),3)` for every thousand pound increase in
weight over the average weight.  For every second a car goes slower in the quarter mile then 
the average seconds per quarter mile, the coefficient for
qsec shows the mpg will increase by `r round(coef(lm2)[3],3)`.  Finally, by using
a manual transmission, the mpg increases `r round(coef(lm2)[4],3)` over the baseline of an
automatic transmission.

Figure 2 in the Appendix shows the regression diagnostic plots for this second model.
The residual plot shows there may be non-linear effect that merit further analysis.

The confidence intervals for the coefficients are
```{r ,echo=FALSE}
confint(lm2)
```

\pagebreak

Appendix
--------
**Summary statistics for Motor Trend Data**
```{r ,echo=FALSE}
summary(mtcars)
```

```{r ,echo=FALSE,fig.width=3.5,fig.height=3.5,fig.cap="MPG Comparision by Transmission Type"}
## determine if mpg is dependent on transmission type
boxplot(cars$mpg~cars$am,xlab="Transmission Type",ylab="mpg")

# means <- tapply(cars$mpg,cars$am,mean)
# points(means,col="red",pch=18,cex=2)
# legend("topleft",c("Diamond represents mean mpg","by Transmission Type"),
#        cex=0.5,
#        box.lty="blank")
```

**Optimal Model Selection**
This documents the series of predictor variables eliminated or added by the **step** function to achieve 
Model 2.  Criteria for model selection is based on the minimum Akaike Information Criteria 
(AIC).  This led to model **`r get.formula(lm2)`**.
```{r ,echo=FALSE}
lm2$anova
```

```{r ,echo=FALSE,warning=FALSE,fig.width=5,fig.height=5,fig.cap="Regression Diagnostics for Model 2"}
par(mfrow=c(2,2),oma=c(0,0,1,0))
plot(lm2)
#mtext("Regression Diagnostics",outer=TRUE)
```

